Equivalence Hypergraphs: E-Graphs for Monoidal Theories

### 2023-11-08 sections/introduction.tex
remove e-rewriting

With consideration of future implementations, we 
additionally provide a combinatorial representation of our enriched string diagrams in terms of (hierarchical)  hypergraphs, which we call \textit{e-hypergraphs}.  We give a specification of e-hypergraph rewriting via a suitable extension of the double pushout (DPO) rewriting framework 
\cite{dpo, bonchi_string_2022-1,bonchi_string_2022-2,bonchi_string_2022}, and prove the representation sound and complete with respect to our categorical semantics.  As a corollary, we also formalize the rewriting theory of standard e-graphs in this way.  Finally, we touch on the question of efficient implementation of rewriting for e-hypergraphs.

@@

\subsection{E-matching and E-rewriting}

Having defined e-hypergraphs and shown them to correctly model the rewriting of string diagrams for semilattice enriched SMCs, we note that the naive implementation of DPO-rewriting is inefficient: to find a redex within an e-hypergraph involves finding a structurally equivalent e-hypergraph which contains the redex as a subgraph. In general, this involves unsharing the e-box structure before a redex becomes available.  Thus, we define a notion of \textit{e-matching} for e-hypergraphs: that is, to find redexes working modulo the e-box structure. In particular, we wish to locate the smallest subgraph $G' \subseteq G$ "containing" (in an appropriate sense) the redex $L$. Given this,  \textit{e-rewriting} $L \to R$, intended to achieve equality saturation, is simple to define due to its non-destructive nature: we rewrite $G' \to G' + R$ in $G$.  
\begin{figure}
\[
	\tikzfig{combinatorial_semantics/example-rewriting}
\]
\caption{Example application of e-rewriting for rewrite $(x*y)/z \to x*(y/z)$: find an appropriate minimal sub-e-hypergraph containing the redex $(x*y)/z$ and non-destructively add the reduct $x*(y/z)$.  REDO THIS}
\end{figure}

>>>>

With consideration of future implementations, we 
additionally provide a combinatorial representation of our enriched string diagrams in terms of (hierarchical)  hypergraphs, which we call \textit{e-hypergraphs}.  We give a specification of e-hypergraph rewriting via a suitable extension of the double pushout (DPO) rewriting framework 
\cite{dpo, bonchi_string_2022-1,bonchi_string_2022-2,bonchi_string_2022}, and prove the representation sound and complete with respect to our categorical semantics.  As a corollary, we also formalize the rewriting theory of standard e-graphs in this way.  %Finally, we touch on the question of efficient implementation of rewriting for e-hypergraphs.

@@

%\subsection{E-matching and E-rewriting}
%
%Having defined e-hypergraphs and shown them to correctly model the rewriting of string diagrams for semilattice enriched SMCs, we note that the naive implementation of DPO-rewriting is inefficient: to find a redex within an e-hypergraph involves finding a structurally equivalent e-hypergraph which contains the redex as a subgraph. In general, this involves unsharing the e-box structure before a redex becomes available.  Thus, we define a notion of \textit{e-matching} for e-hypergraphs: that is, to find redexes working modulo the e-box structure. In particular, we wish to locate the smallest subgraph $G' \subseteq G$ "containing" (in an appropriate sense) the redex $L$. Given this,  \textit{e-rewriting} $L \to R$, intended to achieve equality saturation, is simple to define due to its non-destructive nature: we rewrite $G' \to G' + R$ in $G$.  
%\begin{figure}
%\[
%	\tikzfig{combinatorial_semantics/example-rewriting}
%\]
%\caption{Example application of e-rewriting for rewrite $(x*y)/z \to x*(y/z)$: find an appropriate minimal sub-e-hypergraph containing the redex $(x*y)/z$ and non-destructively add the reduct $x*(y/z)$.  REDO THIS}
%\end{figure}
